---
title: "Interest_debt"
output: html_document
date: "2024-10-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, results = 'hide')
```

```{r}
library(readxl)
library(ggplot2)
library(dplyr)
library(tidyr)
library(png)
library(scales)
library(cowplot)  # For combining plot and logo
library(here)
library(ggalluvial)
```

```{r}
source(here("Reason theme.R"))

logo_path <- "reason_logo.png"

# Load the logo image
logo <- readPNG(here(logo_path))

```

```{r}
# Load the Excel file
outlays <- read_excel("Outlays_functions.xlsx")

#structure
head(outlays)
```


```{r}
# Cleaning the "Function and Subfunction" column by removing numbers and unnecessary spaces
outlays_cleaned <- outlays %>%
  mutate(`Function and Subfunction` = gsub("[0-9]", "", `Function and Subfunction`),
         `Function and Subfunction` = trimws(`Function and Subfunction`))

# Remove "estimate" from column names
outlays_cleaned <- outlays_cleaned %>%
  rename_with(~ gsub(" estimate", "", .), starts_with("202"))

# Preview the cleaned data
head(outlays_cleaned)

```

```{r}
# Filter rows and reshape into long format
outlays_long <- outlays_cleaned %>%
  filter(rowSums(!is.na(select(., matches("^\\d{4}")))) > 0) %>%  # Filter rows with valid data in year columns
  pivot_longer(cols = matches("^\\d{4}"), names_to = "Year", values_to = "Amount") %>%  # Reshape data
  filter(!is.na(Amount)) %>%  # Remove rows with NA in Amount
  mutate(Amount = as.numeric(gsub("[^0-9.]", "", Amount)),  # Clean and convert Amount to numeric
         Year = as.numeric(Year))  # Ensure Year is numeric


```


```{r}
# Filter for Social Security, Medicare, and Interest on Treasury debt securities (gross)
interest_and_social_programs <- outlays_cleaned %>%
  filter(`Function and Subfunction` %in% c("Social security", "Medicare", "Interest on Treasury debt securities (gross)"))

```

```{r}
#Reshape the data into long format (years as a single column)
interest_and_social_programs_long <- interest_and_social_programs %>%
  pivot_longer(cols = -`Function and Subfunction`,  # Select all columns except 'Function and Subfunction'
               names_to = "Year",  # Create a new 'Year' column
               values_to = "Amount")  # Create a new 'Amount' column

# Clean the Amount and Year columns
interest_and_social_programs_long <- interest_and_social_programs_long %>%
  mutate(Amount = as.numeric(gsub("[^0-9.]", "", Amount)),  # Remove any non-numeric characters from Amount
         Year = as.numeric(Year))  # Ensure Year is numeric

# Remove rows with NA Amount values 
interest_and_social_programs_long <- interest_and_social_programs_long %>%
  filter(!is.na(Amount) & !is.na(Year))


```


```{r}
# Create the final plot with the cleaned data
final_plot <- ggdraw(
  ggplot(interest_and_social_programs_long, aes(x = Year, y = Amount, color = `Function and Subfunction`)) +
    geom_line(size = 1.2) +
    labs(
      title = "Rising costs of Social Security, Medicare, and Interest payments are set to overrun the U.S. budget",
      subtitle = "U.S. Federal Outlays Over Time (1990 - 2029)",
      caption = "Source: The Historical Tables at whitehouse.gov",
      x = "Year",
      y = "Outlays (in millions)",
      color = "Category"
    ) +
    theme_minimal() +
    scale_y_continuous(labels = scales::comma) +
    scale_x_continuous(
      breaks = seq(min(interest_and_social_programs_long$Year), 
                   max(interest_and_social_programs_long$Year), by = 5)
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      panel.grid.major.y = element_blank(),
      plot.title = element_text(hjust = 0.5, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5),
      plot.caption = element_text(hjust = 0.5),
      legend.position = "bottom"
    )
) +
  draw_image(logo, x = 0, y = -0.050, width = 0.14, height = 0.14, hjust = 0, vjust = 0)

# Save the final plot with the logo
ggsave("Rising_US_Interest_Payments_vs_Social_Program_Spending_Line_chart.pdf", 
       plot = final_plot, 
       width = 6.2, height = 6)

```

```{r}
# Print the final plot to display it
print(final_plot)
```


Filtering for years starting from 1990 to 2029

```{r}

# Filter data for the three categories we are interested in
filtered_outlays <- outlays_long %>%
  filter(`Function and Subfunction` %in% c("Social security", 
                                           "Medicare", 
                                           "Interest on Treasury debt securities (gross)"))
```

```{r}

# Actual data from 1900 to 2023
actual_data <- filtered_outlays %>%
  filter(Year <= 2023 & Year >= 1900)

# Projected data from 2024 to 2029
projected_data <- filtered_outlays %>%
  filter(Year >= 2024 & Year <= 2029)


```

```{r}
#Create the plot with actual (solid) and projected (dotted) lines
p <- ggplot() +
  
  # Plot actual data with solid lines
  geom_line(data = actual_data, aes(x = Year, y = Amount, color = `Function and Subfunction`), size = 1.2) +
  
  # Plot projected data with dotted lines
  geom_line(data = projected_data, aes(x = Year, y = Amount, color = `Function and Subfunction`), size = 1.2, linetype = "dotted") +
  
  # Add titles and axis labels
  labs(
    title = "U.S. Federal Spending: Actual vs. Projected Costs",
    subtitle = "U.S. Federal Outlays Over Time (1990 - 2029)",
    caption = "Source: The Historical Tables at whitehouse.gov",
    x = "Year",
    y = "Outlays (in millions)",
    color = "Category"
  ) +
  
  # Format the Y-axis with commas
  scale_y_continuous(labels = scales::comma) +
  
  # Add breaks on X-axis every 5 years
  scale_x_continuous(breaks = seq(min(filtered_outlays$Year), max(filtered_outlays$Year), by = 5)) +
  
  # Minimal theme with adjustments
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate X-axis labels
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center title
    plot.subtitle = element_text(hjust = 0.5),  # Center subtitle
    plot.caption = element_text(hjust = 0.5),  # Center caption
    legend.position = "bottom"  # Place legend at the bottom
  ) +
  
  # Annotate forecast period (2024-2029)
  annotate("text", x = 2026, y = 500000, label = "Forecast (2024-2029)", color = "black", size = 4, fontface = "italic")

# Use cowplot to add the logo image to the plot
final_plot <- ggdraw(p) +
  draw_image(logo, x = 0, y = -0.050, width = 0.14, height = 0.14, hjust = 0, vjust = 0)

# Save the final plot with the logo
ggsave(here("Actual_and_projected_Chart.pdf"), plot = final_plot, width = 6.2, height = 6)

# Print the final plot
print(final_plot)



```
##Steady Increase in Social Security and Medicare Costs: The spending on Social Security and Medicare has shown a continuous upward trend, with Social Security having the largest growth in outlays, indicating increasing demands on these social programs.

##Interest Payments Expected to Surge: Projected data from 2024-2029 shows that interest payments on Treasury debt are expected to increase sharply, nearing the level of Medicare spending, signaling a growing financial burden on the federal budget.

#Alluvial Chart 


```{r}
# Create a data frame for actual and projected data combined
combined_data_clean <- rbind(actual_data, projected_data)

# Create rankings for each year based on the Amount (outlays)
combined_data_clean <- combined_data_clean %>%
  group_by(Year) %>%
  mutate(rank = dense_rank(desc(Amount))) %>%
  ungroup()

# Preview the data
head(combined_data_clean)

```


```{r}
# Adjust the combined_data_clean data as needed for ggalluvial
combined_data_clean$Year <- as.factor(combined_data_clean$Year)  # Convert Year to factor for better plotting

# Create the alluvial plot with clearer year intervals and better spacing
p <- ggplot(combined_data_clean,
            aes(axis1 = Year, axis2 = `Function and Subfunction`, y = Amount)) +
  geom_alluvium(aes(fill = `Function and Subfunction`), width = 0.3, knot.pos = 0.3) +  # Adjust the width and knot position
  geom_stratum(width = 0.3, color = "black", size = 0.2) +  # Stratum separation
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 3) +  # Add text labels for categories
  
  # Labels and theme settings
  labs(title = "Tracking U.S. Government Spending Trends: Social Security, Medicare, and Interest Payments",
       y = "Outlays (in millions)",
       x = "Year",
       fill = "Category") +
  theme_minimal() +
  
  # Improve the spacing and clarity of the year axis
  scale_y_continuous(labels = scales::comma) +
  scale_x_discrete(breaks = seq(1990, 2029, by = 5)) +  # Only show year labels every 5 years
  
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for better readability
    panel.grid.major.y = element_blank(),  # Remove horizontal grid lines for a cleaner look
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center-align the title and make it bold
    legend.position = "bottom"  # Move the legend to the bottom for better visibility
  )

# Use cowplot to add the logo image to the plot
final_plot_2 <- ggdraw(p) +
  draw_image(logo, x = 0, y = -0.050, width = 0.14, height = 0.14, hjust = 0, vjust = 0)

# Save the final plot with the logo
ggsave(here("Interest_national_debt_chart_alluvial.pdf"), 
       plot = final_plot, 
       width = 8, height = 8)  # Increase size for better readability


```

```{r}
print(final_plot_2)

```
#The chart shows how the distribution of federal outlays has shifted across these categories, with an increasing share dedicated to Social Security and Medicare, while interest payments on Treasury debt have also been growing steadily over time.

#This visual representation helps highlight:

#Growing Dominance of Social Security and Medicare: As time progresses, more federal resources are allocated to Social Security and Medicare, reflecting the growing demands on these social safety net programs.

#Rising Interest Payments: The chart also captures the increasing burden of interest payments on the federal budget, a key concern for future fiscal sustainability, as seen in the widening red section in the latter years.

```{r}
# Ensure Year is numeric
combined_data_clean$Year <- as.numeric(as.character(combined_data_clean$Year))  # Convert Year to numeric

# Create the stacked area plot including all categories
p <- ggplot(combined_data_clean, aes(x = Year, y = Amount, fill = `Function and Subfunction`)) +
  geom_area(alpha = 0.8, size = 0.5, color = "black") +  # Stacked areas with black borders
  scale_fill_manual(values = c("red", "green", "blue")) +  # Assign colors to categories
  labs(title = "U.S. Government Spending: Social Security, Medicare, and Interest Payments",
       x = "Year",  # Explicitly label as 'Year'
       y = "Outlays (in millions)",
       fill = "Category") +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    panel.grid.major = element_line(color = "gray80"),
    legend.position = "right"
  ) +
  scale_y_continuous(labels = scales::comma) +  # Add commas to Y-axis for better readability
  
  # Correct x-axis: Ensure actual years are displayed
  scale_x_continuous(breaks = seq(1960, 2030, by = 5))  # Set breaks for every 5 years

# Add the logo using cowplot
final_plot_3 <- ggdraw(p) +
  draw_image(logo, x = 0, y = -0.050, width = 0.14, height = 0.14, hjust = 0, vjust = 0)

#  Save the final plot with the logo as a PDF
ggsave("stacked_area_plot_national_debt_chart.pdf", plot = final_plot_3, width = 6.2, height = 6)

# Print the final plot
print(final_plot_3)

```

```{r}

# Plot the data using stacked area plot with facets
p <- ggplot(combined_data_clean, aes(x = Year, y = Amount, fill = `Function and Subfunction`)) +
  geom_area(alpha = 0.8, size = 0.5, color = "black") +  # Stacked area with black borders
  scale_fill_manual(values = c("red", "green", "blue")) +  # Assign colors to categories
  labs(
    title = "faceted_stacked_area_U.S. Government Spending",
    x = "Year",  # Explicitly label X-axis
    y = "Outlays (in millions)",
    fill = "Category"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate X-axis labels for readability
    panel.grid.major = element_line(color = "gray80"),
    legend.position = "right"
  ) +
  scale_y_continuous(labels = scales::comma) +  # Add commas to Y-axis for better readability
  facet_grid(~ `Function and Subfunction`)  # Facet by 'Function and Subfunction'

# Display the plot
print(p)

# Save the plot as a PDF
ggsave("faceted_stacked_area_plot.pdf", plot = p, width = 10, height = 6)



```
```{r}

# Define the groups for plotting
important_groups <- list(
  Defense = c("Military Personnel", "Operation and Maintenance", "Procurement", "Research, Development, Test, and Evaluation", "Subtotal, Department of Defense-Military"),
  International_Affairs = c("International development and humanitarian assistance", "International security assistance", "Conduct of foreign affairs", "Total, International Affairs"),
  Science_Technology = c("General science and basic research", "Space flight, research, and supporting activities", "Total, General Science, Space, and Technology"),
  Health_Social_Security = c("Health care services", "Medicare", "Social security", "Federal employee retirement and disability"),
  Transportation = c("Ground transportation", "Air transportation", "Total, Transportation")
)

# Filter the data for each group
for (group_name in names(important_groups)) {
  group_data <- outlays_long %>%
    filter(`Function and Subfunction` %in% important_groups[[group_name]])
  
  # Create a stacked area plot for each group
  p <- ggplot(group_data, aes(x = Year, y = Amount, fill = `Function and Subfunction`)) +
    geom_area(alpha = 0.8, size = 0.5, color = "black") +  # Stacked areas with black borders
    labs(title = paste(group_name, "U.S. Federal Spending Over Time"),
         x = "Year",
         y = "Outlays (in millions)",
         fill = "Category") +
    theme_reason() +  # Apply the Reason Foundation theme
    theme(
      panel.grid.major.y = element_blank(),  # Remove horizontal grid lines
      plot.title = element_text(hjust = 0.5, face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right"
    ) +
    scale_y_continuous(labels = scales::comma) +  # Format Y-axis with commas
    scale_x_continuous(breaks = seq(min(group_data$Year, na.rm = TRUE), max(group_data$Year, na.rm = TRUE), by = 5))
  
  # Use cowplot to add the logo image to the plot
  final_plot <- ggdraw(p) +
    draw_image(logo, x = 0, y = -0.050, width = 0.14, height = 0.14, hjust = 0, vjust = 0)
  
  # Save each plot with the logo
  ggsave(paste0(group_name, "_spending_with_logo.pdf"), plot = final_plot, width = 10, height = 6)
  
  # Print the plot to display it
  print(final_plot)
}



```

`
